---
title: "Project 1"
author: "Yun Ji"
date: "6/12/2020"
output:
  rmarkdown::github_document:
    toc: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r import, include = FALSE}
library(tidyverse)
library(httr)
library(jsonlite)
```

### Query API Functions
To query the NHL API, I first set up the variables and data structures needed as inputs to my API query functions. The base URL is stored as a character string, and the individual API endpoints are stored in a list called `endpoints` - this is to help future-proof expanding the queries to include other API endpoints later on.
```{r API variables}
baseURL <- "https://records.nhl.com/site/api"

endpoints <- list()
endpoints$franchise <- "/franchise"
endpoints$teamTotals <- "/franchise-team-totals"
endpoints$seasonRecords <- "/franchise-season-records?cayenneExp=franchiseId=ID"
endpoints$goalieRecords <- "/franchise-goalie-records?cayenneExp=franchiseId=ID"
endpoints$skaterRecords <- "/franchise-skater-records?cayenneExp=franchiseId=ID"
```

Next I created functions to support querying the API. The function `createEndpoint` checks the user input for a valid endpoint, and if valid returns the endpoint URL as the output.
```{r createEndpoint function}
createEndpoint <- function(baseURL, endpointList, endpointName) {
  endpoint <- endpointList[[endpointName]]
  
  if(is.null(endpoint)) {
    stop(paste0("Invalid endpoint for ", endpointName))
  } else {
    return(paste0(baseURL, endpoint))
  }
}
```

Functions `getDataNoOptions` and `getDataByFranchiseId` handles the two types of endpoints in the NHL API: those requiring no further input from the user, and those that require the user to specify a numerical value for franchise ID.
```{r getData functions}
getDataNoOptions <- function(fullURL) {
  rawData <- fromJSON(content(GET(fullURL), "text"), flatten=TRUE)
  return(tbl_df(rawData$data))
}

getDataByFranchiseId <- function(fullURL, franchiseId) {
  franchiseId <- as.numeric(franchiseId)
  
  if(is.na(franchiseId)) {
    stop("Franchise ID must be a number")
  } else if(franchiseId < 1 | franchiseId > 38) {
    stop("Franchise ID must between 1 and 38")
  }

  fullURL <- str_replace(fullURL, "ID", as.character(franchiseId))
  rawData <- fromJSON(content(GET(fullURL), "text"), flatten=TRUE)
  return(tbl_df(rawData$data))
}
```

Finally, the wrapper function `runNhlApi` acts as an orchestrator and handles the top-down logic of getting data from the NHL API. The data is returned as a `tibble` table.
```{r runNhlApi function}
runNhlApi <- function(baseURL, endpointList, endpointName, ...) {
  fullURL <- createEndpoint(baseURL, endpointList, endpointName)

  if(endpointName %in% c("franchise","teamTotals")) {
    data <- getDataNoOptions(fullURL)
  } else if(endpointName %in% c("seasonRecords","goalieRecords","skaterRecords")) {
    data <- getDataByFranchiseId(fullURL, ...)
  } else {
    stop("No data found!")
  }
  return(data)
}
```

### Query the Data from the NHL API
I run my functions with the appropriate user inputs to query the needed data from the NHL API. For the franchise-specific data, I picked the Blackhawks since I am from Chicago.
```{r query NHL data}
franchise <- runNhlApi(baseURL, endpoints, "franchise")
teamTotals <- runNhlApi(baseURL, endpoints, "teamTotals")
seasonRecordsBlackhawks <- runNhlApi(baseURL, endpoints, "seasonRecords", 11)
goalieRecordsBlackhawks <- runNhlApi(baseURL, endpoints, "goalieRecords", 11)
skaterRecordsBlackhawks <- runNhlApi(baseURL, endpoints, "skaterRecords", 11)
```

### Plotting Data
For my first plot, I filtered the list of franchises to just the active franchises, then created a scatterplot of their regular season win percentage vs. the starting year of the franchise. I also created the variable `nation` to differentiate Canadian and U.S. teams, and color-coded the data points of the scatterplot by nation.
```{r scatterplot - Win Percent vs. Starting Year}
canada_franchises <- c(1,2,4,5,20,21,25,30,35)

teamTotals <- teamTotals %>%
  mutate(nation = ifelse(franchiseId %in% canada_franchises, "Canada", "USA"))

regularSeasonTeamTotals <- teamTotals %>% 
  filter(gameTypeId == 2, is.na(lastSeasonId)) %>%
  mutate(firstSeason = as.numeric(substr(firstSeasonId,1,4)),
         ties = ifelse(is.na(ties), 0, ties),
         winPercent = (wins+0.5*ties) / gamesPlayed)

g <- ggplot(data = regularSeasonTeamTotals, aes(x = firstSeason, y = winPercent, color = nation))
g + geom_point(size = 4, alpha = 0.5) + 
    labs(title = "NHL Active Teams Regular Season Win Percentage vs. Starting Year",
         x = "Starting Year", y = "Win Percentage",
         color = "Nation", alpha = "") +
    theme(plot.title = element_text(hjust = 0.5)) +
    scale_color_manual(values = c("red","blue"))
```

