---
title: "ST 558 - Summer 2020 - Project 1"
author: "Yun Ji"
date: "6/12/2020"
output:
  rmarkdown::github_document:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r import, include = FALSE}
library(tidyverse)
library(httr)
library(jsonlite)
library(knitr)
```

## Technical Overview

### JSON Data
JavaScript Object Notation, abbreviated as JSON, is a human-readable text object used to store and transmit data. JSON data objects consist of one or more key-value pairs, where the key describes the data field and its value is the data itself. For example, a JSON can contain a key-value pair of `apple: 10` indicating 10 apples and another key-value pair of `orange: 15` for 15 oranges. Key-value pairs in JSON can also be nested, thus imposing hierarchy on the data if needed: an example key of `apples` can have as its value a collection of different types of apples and their counts, such as `red delicious: 8` and `mcintosh: 2`.

JSON data objects are not dependent on any single programming language, and indeed most modern programming languages have functions and packages for creating or reading JSON data. Because of these characteristics, JSON has found wide adoption as a method for transmitting numerical and text data between disparate computer systems and databases.

### JSON Processing in R
In R, there are free, open-source packages to process JSON data objects; these include `rjson`, `RJSONIO` and `jsonlite`. Each package handles JSON slightly differently, including how to process nested structures, missing values, and special characters. `rjson` is generally fastest but has the fewest advanced features. `RJSONIO` has many features but is by far the slowest. `jsonlite` occupies a good middle ground, offering speeds comparable to `rjson` while having nearly as many features as `RJSONIO`, and is therefore my choice to process the JSON data objects involved in this project. Speed and feature comparisons can be found on [this website](https://rstudio-pubs-static.s3.amazonaws.com/31702_9c22e3d1a0c44968a4a1f9656f1800ab.html).

## Query NHL API Functions
To query the NHL API, I first set up the variables and data structures needed as inputs to my API query functions. The base URL is stored as a character string, and the individual API endpoints are stored in a list called `endpoints` - this is to help future-proof expanding the queries in order to add other API endpoints later on.
```{r API variables}
baseURL <- "https://records.nhl.com/site/api"

endpoints <- list()
endpoints$franchise <- "/franchise"
endpoints$teamTotals <- "/franchise-team-totals"
endpoints$seasonRecords <- "/franchise-season-records?cayenneExp=franchiseId=ID"
endpoints$goalieRecords <- "/franchise-goalie-records?cayenneExp=franchiseId=ID"
endpoints$skaterRecords <- "/franchise-skater-records?cayenneExp=franchiseId=ID"
```

Next I created functions to support querying the API. The function `createEndpoint` checks the user input for a valid endpoint, and if valid returns the complete endpoint URL as the output.
```{r createEndpoint function}
createEndpoint <- function(baseURL, endpointList, endpointName) {
  endpoint <- endpointList[[endpointName]]
  
  if(is.null(endpoint)) {
    stop(paste0("Invalid endpoint for ", endpointName))
  } else {
    return(paste0(baseURL, endpoint))
  }
}
```

Functions `getDataNoOptions` and `getDataByFranchiseId` handles the two types of endpoints in the NHL API: those requiring no further input from the user, and those that require the user to specify a numerical value for franchise ID. The functions will query the API for JSON data, flatten the JSON object and convert the data returned into a `tibble` data frame.
```{r getData functions}
getDataNoOptions <- function(fullURL) {
  rawData <- fromJSON(content(GET(fullURL), "text"), flatten = TRUE)
  return(tbl_df(rawData$data))
}

getDataByFranchiseId <- function(fullURL, franchiseId) {
  franchiseId <- as.numeric(franchiseId)
  
  if(is.na(franchiseId)) {
    stop("Franchise ID must be a number")
  } else if(franchiseId < 1 | franchiseId > 38) {
    stop("Franchise ID must between 1 and 38")
  }

  fullURL <- str_replace(fullURL, "ID", as.character(franchiseId))
  rawData <- fromJSON(content(GET(fullURL), "text"), flatten = TRUE)
  return(tbl_df(rawData$data))
}
```

Finally, the wrapper function `runNhlApi` acts as an orchestrator and handles the top-down logic of getting data from the NHL API. Note that unnamed arguments are used in `runNhlApi` to handle the additional user input needed by `getDataByFranchiseId`.
```{r runNhlApi function}
runNhlApi <- function(baseURL, endpointList, endpointName, ...) {
  fullURL <- createEndpoint(baseURL, endpointList, endpointName)

  if(endpointName %in% c("franchise","teamTotals")) {
    data <- getDataNoOptions(fullURL)
  } else if(endpointName %in% c("seasonRecords","goalieRecords","skaterRecords","seasonResults")) {
    data <- getDataByFranchiseId(fullURL, ...)
  } else {
    stop("No data found!")
  }
  return(data)
}
```

## Query Data from the NHL API
I ran my functions with the appropriate user inputs to query the needed data from the NHL API. For the franchise-specific data, I picked the Blackhawks since I am from Chicago.
```{r query NHL data}
franchise <- runNhlApi(baseURL, endpoints, "franchise")
teamTotals <- runNhlApi(baseURL, endpoints, "teamTotals")
seasonRecordsBlackhawks <- runNhlApi(baseURL, endpoints, "seasonRecords", 11)
goalieRecordsBlackhawks <- runNhlApi(baseURL, endpoints, "goalieRecords", 11)
skaterRecordsBlackhawks <- runNhlApi(baseURL, endpoints, "skaterRecords", 11)
```

A sample of each data frame is shown below:
```{r show data frames}
franchise
teamTotals
seasonRecordsBlackhawks
goalieRecordsBlackhawks
skaterRecordsBlackhawks
```

Besides the `id` column which is a row number and therefore not very useful, the other columns have comprehensible names for data (at least to a hockey fan), but some of the data values require additional cleanup, as we will see below. 

## Plotting Data

### History and Winning Percentage of NHL Franchises
For my first plot, I filtered the list of franchises to just the active franchises, performed data wrangling to get a team's first season, fill `NA` values in column `ties` and compute a win percentage using regular season wins, ties, and games played. I then created a scatterplot of their regular season win percentage vs. the starting year of the franchise. I also created the variable `nation` to differentiate Canadian and U.S. teams, and color-coded the data points of the scatterplot by nation.
```{r scatterplot, dpi = 120}
canada_franchises <- c(1,2,4,5,20,21,25,30,35)

teamTotals <- teamTotals %>%
  mutate(nation = ifelse(franchiseId %in% canada_franchises, "Canada", "USA"))

regularSeasonTeamTotals <- teamTotals %>% 
  filter(gameTypeId == 2, is.na(lastSeasonId)) %>%
  mutate(firstSeason = as.numeric(substr(firstSeasonId,1,4)),
         ties = ifelse(is.na(ties), 0, ties),
         winPercent = (wins+0.5*ties) / gamesPlayed)

g <- ggplot(data = regularSeasonTeamTotals, aes(x = firstSeason, y = winPercent, color = nation))
g + geom_point(size = 4, alpha = 0.5) + 
  labs(title = "NHL Active Franchises Regular Season Win Percentage vs. Starting Year",
       x = "Starting Year", y = "Win Percentage",
       color = "Nation", alpha = "") +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(values = c("red","blue"))
```

From the plot, I see that there are many more U.S. teams than Canadian teams, which makes sense as the U.S. is more populous. In addition, there is a big gap in starting year where no active franchises began operations between the early 1930s and the mid 1960s - a long period of separation between the "Original Six" NHL teams and the newer teams added in the NHL's expansion from the late 1960s onwards (there were a few NHL teams that were founded during this interim period, however they ceased operations and are therefore not included among active franchises). Most teams' lifetime winning percentages hover around 50% as one might expect, but there is more variability as one gets closer to the present day.

### Season Results for Central Teams
From the NHL API, I've also queried other endpoints to obtain the season-by-season results of individual teams. To do this, I added a new endpoint to my list of `endpoints`:
```{r new endpoint}
endpoints$seasonResults <- "/franchise-season-results?cayenneExp=franchiseId=ID"
```

I then used this endpoint to get season data for the Chicago Blackhawks as well as for each of Chicago's rival teams in the NHL Central Division. Finally I concatenated them into a single dataframe for the entire Central Divison, then filtered for regular season results only.
```{r get season results}
seasonResultsBlackhawks <- runNhlApi(baseURL, endpoints, "seasonResults", 11)
seasonResultsBlues <- runNhlApi(baseURL, endpoints, "seasonResults", 18)
seasonResultsStars <- runNhlApi(baseURL, endpoints, "seasonResults", 15)
seasonResultsWild <- runNhlApi(baseURL, endpoints, "seasonResults", 37)
seasonResultsPredators <- runNhlApi(baseURL, endpoints, "seasonResults", 34)
seasonResultsJets <- runNhlApi(baseURL, endpoints, "seasonResults", 35)
seasonResultsAvalanche <- runNhlApi(baseURL, endpoints, "seasonResults", 27)

seasonResultsCentral <- rbind(seasonResultsBlackhawks,
                              seasonResultsBlues,
                              seasonResultsStars,
                              seasonResultsWild,
                              seasonResultsPredators,
                              seasonResultsJets,
                              seasonResultsAvalanche)

seasonResultsCentralRegular <- seasonResultsCentral %>%
  filter(gameTypeId == 2) %>%
  mutate(ties = ifelse(is.na(ties), 0, ties),
         winPercent = (wins+0.5*ties) / gamesPlayed)
```

Some data exploration is in order. If a team has won enough games to reach the playoffs in a given season, its playoff results will show up in the `seriesTitle` column of the table (if a team did not make the playoffs, `seriesTitle` is simply `NA`). Using this fact, I created a contingency table showing the counts playoff appearances for each Central Division team since the year 2000.
```{r playoff appearances}
playoffAppearancesSince2000 <- seasonResultsCentralRegular %>% 
  filter(as.numeric(substr(seasonId,1,4)) >= 2000)

cont_table <- table(playoffAppearancesSince2000$teamName, playoffAppearancesSince2000$seriesTitle)
kable(cont_table, caption = "NHL Central Playoff Appearances since 2000")
```

There are a few issues with the above table: the **Atlanta Thrashers** are defunct, having relocated in 2011 as the **Winnipeg Jets**. League changes in 2013 also changed the names of some playoff series: the **Conference Quarterfinals** became the **1st Round** of the NHL playoffs, and the **Conference Semifinals** become the **2nd Round** of the playoffs. Therefore I updated the out-of-date data to their modern-days names, then re-ran the contingency table.
```{r playoff apperances, corrected}
playoffAppearancesSince2000 <- seasonResultsCentralRegular %>% 
  filter(as.numeric(substr(seasonId,1,4)) >= 2000) %>%
  mutate(seriesTitle = ifelse(seriesTitle == "Conference Quarterfinals", "1st Round", seriesTitle),
         seriesTitle = ifelse(seriesTitle == "Conference Semifinals", "2nd Round", seriesTitle),
         teamName = ifelse(teamName == "Atlanta Thrashers", "Winnipeg Jets", teamName))

cont_table <- table(playoffAppearancesSince2000$teamName, playoffAppearancesSince2000$seriesTitle)
kable(cont_table, caption = "NHL Central Playoff Appearances since 2000 (improved)")
```

This table looks better than before. While both the Nashville Predators and St. Louis Blues have reached the playoffs 12 times since 2000, the Chicago Blackhawks took the most advantage of their playoff appearances, reaching (and winning!) the Stanley Cup Final three times in the last 20 years. The Atlanta Thrashers/Winnipeg Jets have experienced the most futility, reaching the playoffs only four times and getting past the 1st round only once.

### Season Results Barplot
Using the `group_by` and `summarise` functions, I aggregated all seasons played by each franchise and tallied the total games played, regular season and playoffs. This includes seasons before year 2000, and required a couple more team name changes for the **Quebec Nordiques** becoming the **Colorado Avalanche** and the **Minnesota North Stars** migrating south as the **Dallas Stars**. Since I don't need `geom_bar` to aggregate my data, I used the `stat = "identity"` option; plotting regular season and playoff bars side-by-side for each franchise instead of stacked required the option `position = "dodge"`. I also used custom bar colors and modified the x-axis franchise names to improve the look of the chart.
```{r season barplot, dpi = 120}
seasonResultsCentralAgg <- seasonResultsCentral %>%
  mutate(teamName = ifelse(teamName == "Atlanta Thrashers", "Winnipeg Jets", teamName),
         teamName = ifelse(teamName == "Minnesota North Stars", "Dallas Stars", teamName),
         teamName = ifelse(teamName == "Quebec Nordiques", "Colorado Avalanche", teamName)) %>%
  group_by(teamName, gameTypeId) %>%
  summarise(totalGamesPlayed = sum(gamesPlayed)) %>%
  mutate(gameType = ifelse(gameTypeId == 2, "Regular Season", "Playoffs"),
         gameType = fct_relevel(gameType, "Regular Season", "Playoffs"))

g <- ggplot(data = seasonResultsCentralAgg, aes(x = teamName, y = totalGamesPlayed, fill = gameType))
g + geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(name = "", values = c("grey40","orange")) +
  scale_x_discrete(labels = c("Chicago Blackhawks" = "Chicago\nBlackhawks",
                              "Colorado Avalanche" = "Colorado\nAvalanche",
                              "Dallas Stars" = "Dallas\nStars",
                              "Minnesota Wild" = "Minnesota\nWild",
                              "Nashville Predators" = "Nashville\nPredators",
                              "St. Louis Blues" = "St. Louis\nBlues",
                              "Winnipeg Jets" = "Winnipeg\nJets")) +
  labs(title = "NHL Central Teams Total Games Played\nRegular Season and Playoffs",
       x = "", y = "Games Played") +
  theme(plot.title = element_text(hjust = 0.5))
```

I spent quite some time getting this chart to display what I intended. I learned that `geom_bar` can be used with a raw data frame to aggregate and visualize some data, but it is not flexible enough to handle all types of aggregations. Sometimes it's simpler to create a new data frame with the exact data you want to present, then use `geom_bar` focusing only on the visualization aspect. And although the barplot is a clear visualization, it doesn't speak to the relative success of these teams: older teams played more regular season games and more playoff games, whereas newer teams had fewer of both.

### Season Results Boxplot
To compare the relative success of Central Division teams, I plotted the win percentage for each season that each team experienced as both a boxplot and a scatterplot. Instead of using `teamName`, this time I utilized `franchiseId` as the team's identifier and renamed this label from the `franchiseId` number to their corresponding team names. Instead of listing teams alphabetically as in the barplot, this time I ordered teams by time of founding to see if in the division there are any trends going from oldest to newest teams.

Note that since the scatterplot data points must come in front of the boxplot, the `geom_point` function is invokved after the `geom_boxplot` function in order to overlay on top of it. With the scatterplot, the `jitter` position option was used to separate the data points across the horizontal axis. Custom colors used to represent each team come from their official team colors, taken from [this helpful website](https://teamcolorcodes.com/nhl-team-color-codes/).
```{r season boxplot, dpi = 120}
g <- ggplot(data=seasonResultsCentralRegular, aes(x = as.factor(franchiseId), y = winPercent, 
                                                  color = as.factor(franchiseId)))
g + geom_boxplot(color = "black", outlier.shape = NA) + 
  geom_point(position = "jitter") + 
  labs(title = "NHL Central Teams Win Percentages for Each Regular Season",
       x = "", y = "Regular Season Win Percentage") + 
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none") +
  scale_x_discrete(labels=c("11" = "Chicago\nBlackhawks",
                            "15" = "Dallas\nStars",
                            "18" = "St. Louis\nBlues",
                            "27" = "Colorado\nAvalanche",
                            "34" = "Nashville\nPredators",
                            "35" = "Winnipeg\nJets",
                            "37" = "Minnesota\nWild")) +
  scale_color_manual(values=c("#CF0A2C","#006847","#002F87","#6F263D","#FFB81C","#041E42","#154734"))
```

With the longest history, the Chicago Blackhawks experienced the widest dispersion of seasonal outcomes and the boxplot reflects this. Newer teams such as the Minnesota Wild and Winnipeg Jets have played fewer seasons, thus have fewer data points and somewhat less dispersion. In the regular season, the Colorado Avalanche and St. Louis Blues have enjoyed the greatest levels of success, with the highest median win percentages.

### Querying Player-Level Data
I also explored player-level data from the `skaterRecords` and `goalieRecords` endpoints in the NHL API. Again, I downloaded data for teams from the Central Division, then concatenated them into a single data frame for use.
```{r query player-level data}
skaterRecordsBlues <- runNhlApi(baseURL, endpoints, "skaterRecords", 18)
skaterRecordsStars <- runNhlApi(baseURL, endpoints, "skaterRecords", 15)
skaterRecordsWild <- runNhlApi(baseURL, endpoints, "skaterRecords", 37)
skaterRecordsPredators <- runNhlApi(baseURL, endpoints, "skaterRecords", 34)
skaterRecordsJets <- runNhlApi(baseURL, endpoints, "skaterRecords", 35)
skaterRecordsAvalanche <- runNhlApi(baseURL, endpoints, "skaterRecords", 27)

skaterRecordsCentral <- rbind(skaterRecordsBlackhawks,
                              skaterRecordsBlues,
                              skaterRecordsStars,
                              skaterRecordsWild,
                              skaterRecordsPredators,
                              skaterRecordsJets,
                              skaterRecordsAvalanche)

goalieRecordsBlues <- runNhlApi(baseURL, endpoints, "goalieRecords", 18)
goalieRecordsStars <- runNhlApi(baseURL, endpoints, "goalieRecords", 15)
goalieRecordsWild <- runNhlApi(baseURL, endpoints, "goalieRecords", 37)
goalieRecordsPredators <- runNhlApi(baseURL, endpoints, "goalieRecords", 34)
goalieRecordsJets <- runNhlApi(baseURL, endpoints, "goalieRecords", 35)
goalieRecordsAvalanche <- runNhlApi(baseURL, endpoints, "goalieRecords", 27)

goalieRecordsCentral <- rbind(goalieRecordsBlackhawks,
                              goalieRecordsBlues,
                              goalieRecordsStars,
                              goalieRecordsWild,
                              goalieRecordsPredators,
                              goalieRecordsJets,
                              goalieRecordsAvalanche)

playerRecordsCentral <- rbind(skaterRecordsCentral %>% select(franchiseName, positionCode), 
                              goalieRecordsCentral %>% select(franchiseName, positionCode))
```

### Displaying Player Data
Since the `positionCode` column codes player positions as single-letter codes, I created an additional field to provide the full names for each position, then displayed the counts of player positions in a contingency table (using `rowSums` to create a new variable for `Total Players`) and a stacked barplot.
```{r player positions, dpi = 120}
playerRecordsCentral <- playerRecordsCentral %>%
  mutate(positionName = "Unknown") %>%
  mutate(positionName = ifelse(positionCode == "C", "Center", positionName),
         positionName = ifelse(positionCode == "D", "Defenseman", positionName),
         positionName = ifelse(positionCode == "L", "Left Wing", positionName),
         positionName = ifelse(positionCode == "R", "Right Wing", positionName),
         positionName = ifelse(positionCode == "G", "Goaltender", positionName))

cont_table <- table(playerRecordsCentral$franchiseName, playerRecordsCentral$positionName)
cont_table <- cbind(cont_table, "Total Players" = rowSums(cont_table))
kable(cont_table, caption = "NHL Central Division Teams - Number of Players by Position")

g <- ggplot(data = playerRecordsCentral, aes(x = franchiseName, fill = positionName))
g + geom_bar() +
  scale_y_continuous(limits = c(0,1000), breaks = seq(0,1200,200), expand = c(0,0)) +
  scale_fill_manual(name = "Player Position",
                    values = c("darkgrey","darkgreen","darkblue","darkred","darkorange")) +
  scale_x_discrete(labels = c("Chicago Blackhawks" = "Chicago\nBlackhawks",
                              "Colorado Avalanche" = "Colorado\nAvalanche",
                              "Dallas Stars" = "Dallas\nStars",
                              "Minnesota Wild" = "Minnesota\nWild",
                              "Nashville Predators" = "Nashville\nPredators",
                              "St. Louis Blues" = "St. Louis\nBlues",
                              "Winnipeg Jets" = "Winnipeg\nJets")) +
  labs(title = "NHL Central Teams: Number of Players at Each Position",
       x = "", y = "Number of Players") +
  theme(plot.title = element_text(hjust = 0.5))
```

Trends here are consistent with other plots in this report: the Blackhawks, with the longest history, have had the greatest number of players with 912. The three youngest teams, the Minnesota Wild, Nasheville Predators and Winnipeg Jets have the lowest player counts. Nasheville's 10 goaltenders represent the lowest count in the table, thanks to the longtime play of goalie Pekka Rinne.

## Blog

This was an interesting project that coalesced many of the data ingestion, exploration, wrangling and visualization methods we have thus far learned in ST 558. The NHL Statistics API was free and easy to use, making it quite a good resource for applying the materials learned. Although it did not have all data (I wanted to make a chart of Stanley Cup winners but could not find the appropriate data), there were plenty to use for a variety of tables and plots.

A difficult part of the project for me was in setting up the functions used to query the API and parse the data. Initially I tried to do everything in one function (not good programming practice) but underestimated its complexity. I then reconsidered, and sketched out on a sheet of paper an outline of key components of the function, before implementing them in R in a methodical way. This enabled me to become better organized in my thinking, and made the programming part easier as well. I will keep up this practice going forward.

Another new experience was developing the project on Github from the RStudio IDE. This was less of a learning curve since I've used Git before, albeit only from a CLI, and the obstacles faced here was mostly setting up Git and syncing it with the IDE. Once up and running, I found the commit/pull/push process fairly easy.

My takeaway from the project is that it got me more practice parsing JSON objects, and quite a lot of hands-on experience with `ggplot2`. I was able to use less common plot features such as custom hexadecimal colors, and in reading through the documentation I realized how much flexibility and user options existed in this one package. I don't have a lot of experience in data visualization, and it's a topic I will be exploring further outside the scope of this course.

## My GitHub Project Repository
https://github.com/yji26/ST558-Project1